---
title: "Exercise 4"
subtitle: Joe Endris
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    df_print: paged
---

```{r libraries, include = FALSE, warning = FALSE, message = FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggfortify)
library(ggtext)
library(grid)
library(gridExtra)
library(tidyverse)
library(lubridate)
library(gtsummary)
library(gt)
library(knitr)
library(flextable)
library(multcomp)
library(multcompView)
library(AICcmodavg)
library(arm)
library(MuMIn)
library(fitdistrplus)
library(pracma)
```

```{r data prep, echo=FALSE, include=FALSE}
#read in phenology observations
phenology<-read_excel("data/phenology_check.xlsx")

#create column for year
phenology <- mutate(phenology, year=year(date))

#create column for julian date
phenology$julian_date <- yday(phenology$date)

#filter out 2021 data since there is no corresponding LT50 data for 2021
phenology <- filter(phenology, year > "2021")

#omit any blank spots in the mean_phenology column
phenology <- phenology[complete.cases(phenology[,4]),]

#calculate mean phenology by julian date
phenology <- phenology%>%
  group_by(species, year, julian_date) %>%
  dplyr::mutate(mean_phenology=mean(phenology))

#calculate SD for phenology
phenology <- phenology%>%
  group_by(species, year, julian_date) %>%
  mutate(pheno_sd = sd(phenology, na.rm=TRUE))

#filter for core three species
phenology <- filter(phenology, species != "Ostrya virginiana")
phenology <- filter(phenology, species != "Quercus alba")

#species as a factor for phenology data
phenology$species <- as.factor(phenology$species)
```
## Background

Spring leaf out is a balancing act.  Too early and a spring cold spell has the risk of damaging photosynthetic tissues.  Too late and a tree's neighbors may have a head start and out compete. 

This is becoming increasingly problematic as winters and springs warm shifting the pheonological norms of temperate species. But as this shift occurs, the mean last frost date is remaining relatively static.

To monitor phenology, a visual observation is conducted on dormant leaf buds and continue on a regular basis through the leaf-out period. For this we used a scale from 0 to 4.

0-	winter dormant
1-	bud swelling/elongating
2-	bud burst
3-	initial leaves
4-	full leaves unfurled

```{r plots, echo=FALSE, include=FALSE, warning = FALSE}
maple_phenology <- ggplot (data=subset(phenology, species=="Acer saccharum"), 
  aes(x = julian_date, y=mean_phenology, color=factor(year))) +
  geom_point()+
  geom_errorbar(aes(y = mean_phenology, ymin = mean_phenology - pheno_sd, ymax = mean_phenology + pheno_sd), alpha = .2) +
  geom_line()+
  labs(x="", y="", colour = "Year")+
  scale_color_manual(values = c("2022" = "blue", "2023" = "red"))+
  ylim(-1, 5)+
  theme_bw()+
  theme(axis.title.x = element_markdown())+
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.position=c("top"),
        text=element_text(size=14))+
  annotate("text", x=40,y=4.5,label= expression(italic("Acer saccharum")), hjust=0, size = 4)

beech_phenology<-ggplot(data=subset(phenology, species=="Fagus grandifolia"), aes(x = julian_date, y=mean_phenology, color=factor(year))) +
  geom_point()+
  geom_errorbar(aes(y = mean_phenology, ymin = mean_phenology - pheno_sd, ymax = mean_phenology + pheno_sd), alpha = .2) +
  geom_line()+
  labs(x="", y="Phenology Code", colour = "Year")+
  scale_color_manual(values = c("2022" = "blue", "2023" = "red"))+
  ylim(-1, 5)+
  theme_bw()+
  theme(legend.position="none")+
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        text=element_text(size=14))+
  annotate("text", x=40,y=4,label= expression(italic("Fagus grandifolia")), hjust=0, size=4)

poplar_phenology<-ggplot(data=subset(phenology, species=="Liriodendron tulipifera"), aes(x = julian_date, y=mean_phenology, color=factor(year))) +
  geom_point()+
  geom_errorbar(aes(y = mean_phenology, ymin = mean_phenology - pheno_sd, ymax = mean_phenology + pheno_sd), alpha = .2) +
  geom_line()+
  labs(x="Julian Date", y="", colour = "Year")+
  scale_color_manual(values = c("2022" = "blue", "2023" = "red"))+
  ylim(-1, 5)+
  theme_bw()+
  theme(legend.position="none")+
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        text=element_text(size=14))+
  annotate("text", x=40,y=4.5,label= expression(italic("Liriodendron tulipifera")), hjust=0, size = 4)

```

```{r phenology plot, warning = FALSE, echo = FALSE}
grid.arrange(maple_phenology, beech_phenology, poplar_phenology, nrow=3)
```
```{r statistical models, include=FALSE, warnings=FALSE, echo = FALSE}
pheno_model <- glm(phenology ~ date + year, data=phenology, family = poisson, na.action="na.fail" )
```

```{r stats summary}
summary(pheno_model)
```
## Plots

Autoplot

```{r autoplot, warnings=FALSE, echo=FALSE}
autoplot(pheno_model)
```

Binned residual plot

```{r residuals plot, echo=FALSE, warnings=FALSE}
x <- predict(pheno_model)
y <- resid(pheno_model)
binnedplot(x, y)
```


## Did you know...
  
    
  
**the Statue of Liberty wears a size 879 shoe!**

<video width="640" height="480" controls>
  <source src="images/gb2.mp4" type="video/mp4">
</video>